Complexity-stability relationships disordered dynamical systems

### 2024-05-06 papers/response-referee.md

- *The mean field equations depend on the expected values of <x_^\gamma> and <x_^{2\gamma}>. The self-consistency condition requires that the probability distribution (12) has exactly the same expected values. This should be discussed in detail.*

We agree with the referee that the discussion about the expected values and self-consistency conditions needed more clarifications.

In some cases, for example $\alpha=1, \beta=3, \gamma=1$, both expectations converge to finite values and we can safely compute the self-consistency equations. However in potential case of interests, and in the example case $\alpha=1, \beta=3/2, \gamma=1$ characterized by a distribution with a power-law tail with exponent $-3/2$, the moments diverge. We now modified the discussion removing the gaussian approximation and considering the introduction of a cut-off. The idea behind this cut-off can be described by the example below.

Consider, as an example which allows for a complete analytical treatment, the case of a power law distribution 
$$
P(x)=\frac{x^{-\alpha}}{\mathcal{Z}} ,
$$
defined from 1 to $\infty$ and with 
$$
\mathcal{Z}=\int_1^{\infty}dxx^{-\alpha}.
$$

Consider the case $\alpha=3/2$. The distribution is normalized with $\mathcal{Z}=2$, but the mean diverges. However, we would like to be able to describe the behaviour of the sample mean 
$$
\bar{x}_N\equiv\frac{1}{N}\sum_{i=1}^Nx_i,
$$
where the $x_i$ are extracted from $P(x)$.
For this purpose, we can define the quantity
$$
\langle x\rangle_{\Lambda}\equiv\int_1^{\Lambda}dxP(x)x,
$$
with the cut-off $\Lambda$ defined such that $\int_{\Lambda}^{\infty}dxP(x)=1/N$, i.e., such that there is statistically less than 1 variable with value above $\Lambda$ out of $N$ extracted variables. For the case $\alpha=3/2$ we have 
$$
\frac{1}{2}\int_{\Lambda}^{\infty}dxx^{-3/2}=\Lambda^{-1/2},
$$
and therefore $\Lambda=N^2$. We have for the mean
$$
\langle x\rangle_{\Lambda}=\frac{1}{2}\int_1^{N^2}dxx^{-1/2} = N-1.
$$
The result is plotted below, alongside sample mean for extractions of $N=10,$ $10^2,$ $10^3,$ $10^4,$ $10^5,$ $10^6$.

### 2024-05-06 papers/submission/beyond-may.tex

We compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = - A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) - A_{ii}g(x_i^*)h'(x_j^*) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}A_{ij}h(x_j^*)=f(x_i^*)/g(x_i^*)$.
In order to investigate the spectral properties of $J^*$, 
we follow Stone~\cite{Stone2018} in using a recent generalization of the circular law in random matrix theory~\cite{Ahmadian2015}.

@@

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative radnom variable $x_*$
\begin{equation}
    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\tilde{\mu}=\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\tilde{\sigma}^2 \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \tilde{\mu} \langle x_*^{\gamma}\rangle)^2}{2\tilde{\sigma}^2\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently (see~\cite{Hatton2024} for a discussion on feasibility for systems described by this kind of distributions).

@@

%Conclusion
\paragraph*{}
\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
In this letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that the competitive Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where self-interactions grow faster than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.

>>>>

We compute the Jacobian matrix at equilibrium
\begin{align}
    J_{ij}^* & = - A_{ij}g(x_i^*)h'(x_j^*) \qquad \qquad \textrm{for} \ i\neq j \label{eq: jac off-diag}\\
    J_{ii}^* & = f'(x_i^*) - g'(x_i^*)f(x_i^*)/g(x_i^*) - A_{ii}g(x_i^*)h'(x_i^*) \ , \label{eq: jac diag}
\end{align}
where we used $\sum_{j}A_{ij}h(x_j^*)=f(x_i^*)/g(x_i^*)$.
In order to investigate the spectral properties of $J^*$, 
we follow Stone~\cite{Stone2018} in using a recent generalization of the circular law in random matrix theory~\cite{Ahmadian2015}.

@@

%Random interactions: equilibrium distribution
\paragraph*{}
\emph{Random interactions: equilibrium distribution.---}Henceforth we assume $f(x_i)=x_i^{\alpha}$, $g(x_i)=x_i^{\beta}$, $h(x_i)=x_i^{\gamma}$. 
(Any coefficients can be reabsorbed in the statistics of $A$ and by a rescaling of time.)
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{beta1_5-S50-N10-diversity-increase.pdf}
    \caption{For a system in the ``anti-May" phase, an increase in $N$,
    corresponding to moving along $\sqrt{N}$ in the $(\sigma \sqrt{N},\mu N)$ plane (green dashed line), will be stabilizing rather than destabilizing.
    Here we show results from the numerical resolution of the dynamical system~\eqref{dynamics}. Stability is defined as full stable coexistence. The red line is computed using the cavity solution and the generalized stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
    Parameters values are $\alpha=1$, $\beta=3/2$,
    $\gamma=1$, $N=50$ and $10$ replicates for the simulations. The green line is plotted for $\mu=0.1$ and $\sigma=0.75$.}
    \label{fig: stability line + sims}
\end{figure}
Following e.g. Ref.~\cite{Roy2019}, we can derive from Eq.~\eqref{dynamics} a cavity solution which describes the ensemble of the stationary value of $N\gg 1$ variables by means of a single representative random variable $x_*$
\begin{equation}
    0 = x_*^{\alpha}-A_sx_*^{\beta+\gamma}-x_*^{\beta}\big( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\big) \, ,
\label{eq: dmft}
\end{equation}
where $A_s$ is a random variable with the statistics of $A_{ii}$ and $\xi$ is a standard normal random variable. 
\begin{figure}[t!]
    \includegraphics[width=.45\textwidth]{alpha-beta.pdf}
    \caption{The ``May" and ``anti-May" phases derived in the
    case of uniform interactions ($\sigma = 0$) are robust with respect to random interactions ($\sigma > 0$).
    Here we simulated system~\eqref{dynamics} in the power law case for a fixed value of
    $\mu=\mu_s=10^{-2}$, $\gamma=1$, $N=100$ and $100$ replicates,  
    and we evaluated $\sigma_c$, defined here as
    the value above which full stable coexistence is less probable than 50\%,
    at varying $\alpha$ and $\beta$.
    No amount of heterogeneity is allowed in the bottom right triangle because we have set $\mu=\mu_s$, while an increasing value of $\sigma_c$
    is found as $\beta$ becomes larger than $\alpha$.
    We stress that the existence of a finite $\sigma_c\sqrt{N}$ above which instability is triggered also in the upper triangle,
    does not imply that increasing $N$, at fixed $\mu$ and $\sigma$, will destabilize the system, because $\tilde{\mu}=\mu N$ would also change.
    }
    \label{fig: alpha-beta}
\end{figure}
The equation above can be solved for $x_*$ for specific values of
$(\alpha-\beta)/\gamma$, or if $A_s=0$.
(In the following we show that neglecting $A_s$, when it is of the order of $\mu$, does not affect noticeably the quality of the approximation.) In this case the stationary solution is given by 
\begin{equation} \label{eq: cavity solution}
    x_* = \left( \tilde{\mu} \langle x_*^{\gamma}\rangle + \tilde{\sigma} \sqrt{\langle x_*^{2\gamma}\rangle}\xi\right)^{1/(\alpha-\beta)} \, .
\end{equation}
The equilibrium distribution $P(x_*)$ can then be obtained as the pushforward of the distribution of $\xi$: 
\begin{equation}\label{eq: dist general}
    P(x_*)=\frac{|\alpha-\beta|x_*^{\alpha-\beta-1}}{\sqrt{2\pi\tilde{\sigma}^2 \langle x_*^{2\gamma}\rangle}}
    \exp{\left\{-\frac{(x_*^{\alpha-\beta}- \tilde{\mu} \langle x_*^{\gamma}\rangle)^2}{2\tilde{\sigma}^2\langle x_*^{2\gamma}\rangle}\right\}} \, ,
\end{equation}
defined from $0$  to $\infty$, and where the expectation values must be computed self-consistently. XXX feasibility XXX

@@

%Conclusion
\paragraph*{}
\emph{Discussion.---}The relationship between complexity and stability in high-dimensional dynamical systems has been a longstanding puzzle, in ecology and other fields. 
In this letter, we have shown that the condition $\sigma\sqrt{N}< \mu_s$ does not provide a complete picture of the relationship between complexity and stability. 
In particular, we have seen that the competitive Lotka-Volterra model, often cited in support of May's general prediction, exemplifies a special cancellation in the more general stability condition $N\langle (\sigma/\psi)^2\rangle < 1$.
In models where self-interactions grow slower than cross-interactions~\cite{Hatton2024,samadder2024interconnection}, the opposite behavior is observed: stability becomes \emph{more} likely with increasing diversity $N$.
